---
title: "05_parent_bleaching_stats"
output: html_document
date: "2024-08-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##col statistics
```{r}
set.seed(123)
library(tidyverse) #for data wrangling
library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(ggfortify) #for model diagnostics
library(sjPlot)    #for outputs
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(MuMIn)     #for AICc
library(brms)
library(broom.mixed)
library(tidybayes)
library(bayesplot)
#library(standist)   #for visualizing distributions
library(rstanarm)
library(ggeffects)
library(rstan)
library(DHARMa)
library(ggridges)
library(easystats)     #framework for stats, modelling and visualisation
library(patchwork)
source('/Users/alexmacadam/Documents/GitHub/AGF-Outplanting/helperFunctions.R')
```

```{r}
# Function for summarising col standard error
summarySE <- function(data=NULL, measurevar=NULL, groupvars=NULL, na.rm=FALSE,
                      conf.interval=.95, .drop=TRUE) {
  require(plyr)
  
  # New version of length which can handle NA's: if na.rm==T, don't count them
  length2 <- function (x, na.rm=FALSE) {
    if (na.rm==TRUE) sum(!is.na(x))
    else       length(x)
  }
  
  # This does the summary. For each group's data frame, return a vector with
  # N, mean, and sd
  datac <- ddply(data, groupvars, .drop=.drop,
                 .fun = function(xx, col) {
                   c(N    = length2(xx[[col]], na.rm=na.rm),
                     mean = mean   (xx[[col]], na.rm=na.rm),
                     sd   = sd     (xx[[col]], na.rm=na.rm)
                   )
                 },
                 measurevar
  )
  
  # Commenting this rename mean column line, as not assigning measurevar name properly
  # # Rename the "mean" column    
  # datac <- rename(datac, measurevar = mean)
  
  datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean
  
  # Confidence interval multiplier for standard error
  # Calculate t-statistic for confidence interval: 
  # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
  ciMult <- qt(conf.interval/2 + .5, datac$N-1)
  datac$ci <- datac$se * ciMult
  
  return(datac)
}
```

```{r}
#load data
colour <- read.csv("/Users/alexmacadam/Documents/GitHub/AGF-Outplanting/data/adult_colour.csv", header=T, fileEncoding="UTF-8-BOM") |> filter(Temp == "Hot")

colour <- colour |>
  dplyr::mutate(
         Reef = factor(Reef, levels = c('Davies', 'Palms', 'Arlington', 'Martin', 'Wood')),
         Tank = factor(Tank),
         Temp = factor(Temp)
         ) |>
  dplyr::filter(Timepoint != "1") |>
  dplyr::filter(Species == "A. hyacinthus")
```

```{r}
colour$reefxtemp <- paste(colour$Reef, colour$Temp, sep='_')
colour$reefxtemp <- as.factor(colour$reefxtemp)
```

Colour summary
```{r}
col_sum2 <- colour |>
  dplyr::filter(Timepoint == "4") |>
  summarySE(measurevar="CalibratedMeanGrey",
                     groupvars=c("Reef", "Temp"), na.rm=TRUE)
```

Colour change summary
```{r}
bleaching<- colour |>
  tidyr::pivot_wider(names_from = "Timepoint",
    values_from = "CalibratedMeanGrey",
    id_cols = c("CoralID","Tank", "Temp", "Reef")) |>
  dplyr::mutate(Bleaching = `4` - `0`)

bleach_sum <- bleaching |>
  summarySE(measurevar="Bleaching",
                     groupvars=c("Reef", "Temp"), na.rm=TRUE)
```

```{r}
bleaching$reefxtemp <- paste(bleaching$Reef, bleaching$Temp, sep='_')
bleaching$reefxtemp <- as.factor(bleaching$reefxtemp)
```


##############################################
Colour Stats
# Fit the model
## Random intercept
form
```{r}
col.form <- bf(CalibratedMeanGrey ~ Reef + (1|Tank),
                     family = gaussian()
                   )
```

```{r}
options(width=150)
col.form %>% get_prior(data = colour)
options(width=80)
```

priors
```{r fitModel2h1, results='markdown', eval=TRUE, mhidden=TRUE, cache=TRUE}
priors<- prior(normal(3, 2.5), class = "Intercept") +
    prior(normal(0, 2), class = "b") +
    prior(normal(0, 2), class = "sd", group = "Tank") +
    #prior(normal(0, 2), class = "sd", group = "Temp") +
    prior(student_t(3, 0, 2), class = 'sigma') +
    prior(cauchy(0,2), class = 'sd')
```

run model- n
```{r}
col.brm2 <- brms::brm(col.form, 
                  data = colour,
                  prior = priors,
                  sample_prior = 'yes',
                  iter = 5000,
                  warmup = 2500,
                  chains = 3, cores = 3,
                  thin = 5,
                  control = list(adapt_delta = 0.99, max_treedepth = 20),
                  refresh = 0,
                  backend = "rstan"
                  )
```

```{r}
#saveRDS(col.brm2, file = "data/adult.col.brm2.rds")
col.brm2 <- readRDS("data/adult.col.brm2.rds")
```

Assess model
```{r}
col.brm2 %>% SUYR_prior_and_posterior()
```

```{r}
col.brm2$fit |> stan_trace()
```

```{r}
col.brm2$fit |> stan_ac()
```

```{r}
col.brm2$fit |> stan_rhat()
```

```{r}
col.brm2$fit |> stan_ess()
```

```{r}
col.brm2 |> pp_check(type = 'dens_overlay', nsamples = 250)
```

Dharma residuals
```{r}
col.resids<- make_brms_dharma_res(col.brm2, integerResponse = FALSE)
wrap_elements(~testUniformity(col.resids)) +
wrap_elements(~plotResiduals(col.resids, form = factor(rep(1,nrow(col))))) +
wrap_elements(~plotResiduals(col.resids, quantreg = FALSE))+
wrap_elements(~testDispersion(col.resids))
```

conditional effects
```{r}
col.brm2 %>%
    conditional_effects() %>%
    plot(points = TRUE)
```

Summarise model
back transform from logit scale: by exponentiating on the odds scale. comparing to 1
```{r}
col.brm2 |>
  as_draws_df() |>
  mutate(across(everything(), exp)) |>
  summarise_draws(
    median,
    HDInterval::hdi,
    Pl = ~mean(.x < 0),
    Pg = ~mean(.x > 0),
    rhat,
    ess_bulk,
    ess_tail
    ) |>
  knitr::kable()
```

Probability scale
```{r}
col.brm2 |>
  emmeans(~Reef, type = "response")
```

Planned contrast:
Davies
Palms
Arlington
Martin
Wood

```{r}
cmat=cbind(
  "North vs South" = c(0, -1/3, 0, -1/3, 0, 1/2, 0, -1/3, 0, 1/2)
)

cmat=cbind(
  "North vs South" = c(-1/3, -1/3, -1/3, 1/2, 1/2)
)
```

```{r}
col.em<- col.brm2 |>
  emmeans(~Reef, type = 'response') |>
  contrast(method=list(cmat)) |>
  gather_emmeans_draws() |>
  mutate(across(everything(), exp)) |>
#on odds ratio scale
  dplyr::summarise(median_hdci(.value),
            Pl = mean(.value < 0),
            Pg = mean(.value > 0)
            )
col.em
```

```{r}
col.plot<- col.brm2 %>%
  emmeans(~Reef) %>%
  gather_emmeans_draws() %>%
    ggplot() +
  #geom_vline(xintercept = 0, linetype = "dashed") +
  stat_slab(aes(
    x = .value, y = Reef,
    fill = stat(ggdist::cut_cdf_qi(cdf,
      .width = c(0.5, 0.8, 0.95),
      labels = scales::percent_format()
    ))
  ), color = "black") +
  scale_fill_brewer("Interval", direction = -1, na.translate = FALSE) +
  theme_classic() +
  labs(x = "col", y = "Reef")
col.plot
```

```{r}
#saveRDS(col.plot, file = "data/adult.col.plot.rds")
#col.plot <- readRDS("data/adult.col.plot.rds")
```

##############################################
Bleaching Stats
# Fit the model
form
```{r}
bleach.form <- bf(Bleaching ~ Reef + (1|Tank),
                     family = gaussian()
                   )
```

priors
```{r fitModel2h1, results='markdown', eval=TRUE, mhidden=TRUE, cache=TRUE}
priors<- prior(normal(3, 2.5), class = "Intercept") +
    prior(normal(0, 2), class = "b") +
    prior(normal(0, 2), class = "sd", group = "Tank") +
    #prior(normal(0, 2), class = "sd", group = "Temp") +
    prior(student_t(3, 0, 2), class = 'sigma') +
    prior(cauchy(0,2), class = 'sd')
```

run model- n
```{r}
bleach.brm2 <- brms::brm(bleach.form, 
                  data = bleaching,
                  prior = priors,
                  sample_prior = 'yes',
                  iter = 5000,
                  warmup = 2500,
                  chains = 3, cores = 3,
                  thin = 5,
                  control = list(adapt_delta = 0.99, max_treedepth = 20),
                  refresh = 0,
                  backend = "rstan"
                  )
```

```{r}
#saveRDS(bleach.brm2, file = "data/adult.bleach.brm2.rds")
bleach.brm2 <- readRDS("data/adult.bleach.brm2.rds")
```

Assess model
```{r}
bleach.brm2 %>% SUYR_prior_and_posterior()
```

```{r}
bleach.brm2$fit |> stan_trace()
```

```{r}
bleach.brm2$fit |> stan_ac()
```

```{r}
bleach.brm2$fit |> stan_rhat()
```

```{r}
bleach.brm2$fit |> stan_ess()
```

```{r}
bleach.brm2 |> pp_check(type = 'dens_overlay', nsamples = 250)
```

Dharma residuals
```{r}
bleach.resids<- make_brms_dharma_res(bleach.brm2, integerResponse = FALSE)
wrap_elements(~testUniformity(bleach.resids)) +
wrap_elements(~plotResiduals(bleach.resids, form = factor(rep(1,nrow(bleach))))) +
wrap_elements(~plotResiduals(bleach.resids, quantreg = FALSE))+
wrap_elements(~testDispersion(bleach.resids))
```

conditional effects
```{r}
bleach.brm2 %>%
    conditional_effects() %>%
    plot(points = TRUE)
```

Summarise model
back transform from logit scale: by exponentiating on the odds scale. comparing to 1
```{r}
bleach.brm2 |>
  as_draws_df() |>
  mutate(across(everything(), exp)) |>
  summarise_draws(
    median,
    HDInterval::hdi,
    Pl = ~mean(.x < 1),
    Pg = ~mean(.x > 1),
    rhat,
    ess_bulk,
    ess_tail
    ) |>
  knitr::kable()
```

Probability scale
```{r}
bleach.brm2 |>
  emmeans(~Reef, type = "response")
```

Planned contrast:
 Arlington_Ambient
 Arlington_Hot
 Davies_Ambient
 Davies_Hot
 Martin_Ambient
 Martin_Hot
 Palms_Ambient
 Palms_Hot
 Wood_Ambient
 Wood_Hot

```{r}
cmat=cbind(
  "North vs South" = c(-1/3, -1/3, 1/2, -1/3, 1/2)
)
```

```{r}
bleach.em<- bleach.brm2 |>
  emmeans(~Reef, type = 'response') |>
  contrast(method=list(cmat)) |>
  gather_emmeans_draws() |>
  mutate(across(everything(), exp)) |>
#on odds ratio scale
  dplyr::summarise(median_hdci(.value),
            Pl = mean(.value < 0),
            Pg = mean(.value > 0)
            )
bleach.em
```

```{r}
bleach.plot<- bleach.brm2 %>%
  emmeans(~Reef) %>%
  gather_emmeans_draws() %>%
    ggplot() +
  #geom_vline(xintercept = 0, linetype = "dashed") +
  stat_slab(aes(
    x = .value, y = Reef,
    fill = stat(ggdist::cut_cdf_qi(cdf,
      .width = c(0.5, 0.8, 0.95),
      labels = scales::percent_format()
    ))
  ), color = "black") +
  scale_fill_brewer("Interval", direction = -1, na.translate = FALSE) +
  theme_classic() +
  labs(x = "bleach", y = "Reef")
bleach.plot
```

```{r}
#saveRDS(bleach.plot, file = "data/adult.bleach.plot.2.rds")
#bleach.plot <- readRDS("data/adult.bleach.plot.2.rds")
```
