---
title: "07_juvenile_size_stats"
author: "Alex Macadam"
date: "2023-10-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##growth statistics
```{r}
library(tidyverse) #for data wrangling
library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(ggfortify) #for model diagnostics
library(sjPlot)    #for outputs
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(MuMIn)     #for AICc
library(brms)
library(broom.mixed)
library(tidybayes)
library(bayesplot)
#library(standist)   #for visualizing distributions
library(rstanarm)
library(ggeffects)
library(rstan)
library(DHARMa)
library(ggridges)
library(easystats)     #framework for stats, modelling and visualisation
library(patchwork)
source('/Users/alexmacadam/Documents/GitHub/AGF-Outplanting/helperFunctions.R')
```

```{r}
#load data
size <- read.csv("/Users/alexmacadam/Documents/GitHub/AGF-Outplanting/data/juvenile_size.csv", header=T, fileEncoding="UTF-8-BOM")
```

```{r}
#change catagorical columns to factors and remove NAs
size <- size |>
  mutate(Family = factor(Family),
         Timepoint = factor(Timepoint),
         Cone = factor(Cone),
         Cross = factor(Cross),
         popcross = factor(popcross, levels = c('PIxPI', 'DAxDA', 'DAxAR', 'DAxMA', 'MAxDA', 'MAxMA', 'WDxMA')),
         RegionCross = factor(RegionCross),
         Breed = factor(Breed),
         Rack = factor(Rack),
         DeviceNo = factor(DeviceNo)
         ) |>
  filter(Area != "NA")
```

Check all entries are unique
```{r}
size %>% filter(Timepoint == "T1") |>
  dplyr::group_by(Cross, Rack, DevicePos, DeviceNo, IndID, popcross) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::filter(n > 1L)
```

```{r}
growth<- size %>% filter(Timepoint %in% c("T0","T1")) %>% dplyr::select(Family, Timepoint, popcross, Area, IndID, Cone, RegionCross, Rack, DeviceNo) %>%
  pivot_wider(names_from = Timepoint, values_from = Area) %>%
  tidyr::unnest() %>%
  mutate(Growth = ((`T1` - `T0`)/`T0`)*100) %>%
  drop_na(Growth)
```

# Fit the model
## Random intercept
form
```{r}
growth.form <- bf(Growth ~ popcross + (1|Rack),
                     family = gaussian()
                   )
```

```{r}
growth |>
group_by(popcross) |>
summarise(
mean(Growth),
median(Growth),
sd(Growth),
mad(Growth)
)
```

priors
```{r fitModel2h1, results='markdown', eval=TRUE, mhidden=TRUE, cache=TRUE}
priors<- prior("normal(250, 100)", class = "Intercept") +
    prior("normal(0, 200)", class = "b") +
    prior("normal(0, 130)", class = "sd", group = "Rack") +
   # prior("normal(0, 130)", class = "sd", group = "Timepoint") +
    prior(student_t(3, 0, 140), class = 'sigma') +
    prior(cauchy(0,2.5), class = 'sd')
```

run model
```{r}
growth.brm2 <- brm(growth.form, 
                  data = growth,
                  prior = priors,
                  sample_prior = 'yes',
                  iter = 5000,
                  warmup = 2500,
                  chains = 3, cores = 3,
                  thin = 5,
                  control = list(adapt_delta = 0.99, max_treedepth = 20),
                  refresh = 0,
                  backend = "rstan"
                  )
```

```{r}
#saveRDS(growth.brm2, file = "data/juv.growth.brm2.rds")
growth.brm2 <- readRDS("data/juv.growth.brm2.rds")
```

conditional effects
```{r}
growth.brm2 %>%
    conditional_effects("popcross") %>%
    plot(points = TRUE)
```

```{r}
growth.brm2 %>% SUYR_prior_and_posterior()
```

```{r}
growth.brm2$fit |> stan_trace()
```

```{r}
growth.brm2$fit |> stan_ac()
```

```{r}
growth.brm2$fit |> stan_rhat()
```

```{r}
growth.brm2$fit |> stan_ess()
```

```{r}
growth.brm2 |> pp_check(type = 'dens_overlay', nsamples = 250)
```

Dharma residuals
```{r}
growth.resids<- make_brms_dharma_res(growth.brm2, integerResponse = FALSE)
wrap_elements(~testUniformity(growth.resids)) +
wrap_elements(~plotResiduals(growth.resids, form = factor(rep(1,nrow(growth))))) +
wrap_elements(~plotResiduals(growth.resids, quantreg = FALSE))+
wrap_elements(~testDispersion(growth.resids))

testUniformity(growth.resids)
plotResiduals(growth.resids, form = factor(rep(1,nrow(growth))))
plotResiduals(growth.resids, quantreg = FALSE)
testDispersion(growth.resids)
```

Summarise model
back transform from logit scale: by exponentiating on the odds scale. comparing to 1
```{r}
growth.brm2 |>
  summarise_draws(
    median,
    HDInterval::hdi,
    Pl = ~mean(.x < 0),
    Pg = ~mean(.x > 0),
    rhat,
    ess_bulk
    ) |>
  knitr::kable()
```

Probability scale
```{r}
growth.brm2 |>
  emmeans(~popcross) |>
  pairs()
```

Planned contrast:
PIxPI
DAxDA 
DAxAR     
DAxMA              
MAxDA         
MAxMA              
WDxMA   

```{r}
cmat=cbind(
  "Pure vs Hybrid" = c(-1/3, -1/3, 1/4, 1/4, 1/4, -1/3, 1/4),
  "PureCentral vs Hybrid" = c(-1/2, -1/2, 1/3, 1/3, 1/3, 0, 0),
  "PureCentral vs Hybrid 2" = c(-1/2, -1/2, 1/4, 1/4, 1/4, 0, 1/4),
  "PureNorth vs Hybrid" = c(0, 0, 1/3, 1/3, 1/3, -1, 0),
  "PureNorth vs Hybrid2" = c(0, 0, 1/3, 1/3, 1/3, -1/2, -1/2),
  "Palms vs Davies" = c(1, -1, 0, 0, 0, 0, 0),
  "MARmom vs MARdad" = c(0, 0, 0, 1, -1, 0, 0),
  "PureCentral vs PureNorth" = c(-1/2, -1/2, 0, 0, 0, 1/2, 1/2),
  "PureCentral vs PureNorth2" = c(-1/2, -1/2, 0, 0, 0, 1, 0),
  "PIxPI vs hybrids" = c(-1, 0, 1/3, 1/3, 1/3, 0, 0)
)
```


odds scale:
```{r}
growth.em<- growth.brm2 |>
  emmeans(~popcross) |>
  contrast(method=list(cmat)) |>
  gather_emmeans_draws() |>
#on odds ratio scale
  summarise(median_hdci(.value),
            Pl = mean(.value < 0),
            Pg = mean(.value > 0)
            )
growth.em
```
  "Pure vs Hybrid": y = 10.538. There is no evidence that Hybrids have higher growth rate than pure juveniles (Pg = 0.859)
  
  "PureCentral vs Hybrid": y = -5.139. There is no evidence that PureCentrals have higher growth rate than hybrid juveniles (Pl = 0.664)
  "PureCentral vs Hybrid 2": y = 1.392. There is no evidence that PureCentrals have higher growth rate than hybrid juveniles (Pg = 0.559) 
  
  "PureNorth vs Hybrid": y = 23.096. Hybrids have 23% higher growth rate than pure north juveniles (Pg = 0.966)
  "PureNorth vs Hybrid2": y = -1.278. There is no evidence that hybrids have higher growth rate than PureNorth juveniles (Pl = 0.531)
  
  "Palms vs Davies": y = -46.838. Palms has 46.8% higher growth rate than Davies (Pl = 1.000)
  
  "PureCentral vs PureNorth": y = -4.008. There is no evidence that PureCentrals have higher growth rate than PureNorth juveniless (Pl = 0.653)
  "PureCentral vs PureNorth2": y = -27.811. PureCentral have 27.8% higher growth rate than PureNorth juveniles (Pg = 0.999)
  
  "MARmom vs MARdad": y = -10.130. There is no evidence that Hybrids have higher growth rate than pure juveniles (Pg = 0.645)
  
#Plot
get variables
```{r}
growth.brm2 %>% get_variables()
```

```{r}
growth.plot<- growth.brm2 %>%
  emmeans(~popcross) %>%
  gather_emmeans_draws() %>%
    ggplot() +
  #geom_vline(xintercept = 0, linetype = "dashed") +
  stat_slab(aes(
    x = .value, y = popcross,
    fill = stat(ggdist::cut_cdf_qi(cdf,
      .width = c(0.5, 0.8, 0.95),
      labels = scales::percent_format()
    ))
  ), color = "black") +
  scale_fill_brewer("Interval", direction = -1, na.translate = FALSE) +
  theme_classic()+
  labs(x = "Growth (%)", y = "Population Cross")
growth.plot
```

```{r}
#saveRDS(growth.plot, file = "data/juv.growth.plot.reef.rds")
growth.plot <- readRDS("data/juv.growth.plot.reef.rds")
```

##############
# Fit the model
## Random intercept
form
```{r}
growth.form <- bf(Growth ~ RegionCross + (1|Rack),
                     family = gaussian()
                   )
```

```{r}
growth |>
group_by(popcross) |>
summarise(
mean(Growth),
median(Growth),
sd(Growth),
mad(Growth)
)
```

priors
```{r fitModel2h1, results='markdown', eval=TRUE, mhidden=TRUE, cache=TRUE}
priors<- prior("normal(250, 100)", class = "Intercept") +
    prior("normal(0, 200)", class = "b") +
    prior("normal(0, 130)", class = "sd", group = "Rack") +
   # prior("normal(0, 130)", class = "sd", group = "Timepoint") +
    prior(student_t(3, 0, 140), class = 'sigma') +
    prior(cauchy(0,2.5), class = 'sd')
```

run model
```{r}
growth.brm3 <- brm(growth.form, 
                  data = growth,
                  prior = priors,
                  sample_prior = 'yes',
                  iter = 5000,
                  warmup = 2500,
                  chains = 3, cores = 3,
                  thin = 5,
                  control = list(adapt_delta = 0.99, max_treedepth = 20),
                  refresh = 0,
                  backend = "rstan"
                  )
```

```{r}
#saveRDS(growth.brm3, file = "data/juv.growth.brm3.rds")
#growth.brm3 <- readRDS("data/juv.growth.brm3.rds")
```

conditional effects
```{r}
growth.brm3 %>%
    conditional_effects("popcross") %>%
    plot(points = TRUE)
```

```{r}
growth.brm3 %>% SUYR_prior_and_posterior()
```

```{r}
growth.brm3$fit |> stan_trace()
```

```{r}
growth.brm3$fit |> stan_ac()
```

```{r}
growth.brm3$fit |> stan_rhat()
```

```{r}
growth.brm3$fit |> stan_ess()
```

```{r}
growth.brm3 |> pp_check(type = 'dens_overlay', nsamples = 250)
```

Dharma residuals
```{r}
growth.resids<- make_brms_dharma_res(growth.brm3, integerResponse = FALSE)
wrap_elements(~testUniformity(growth.resids)) +
wrap_elements(~plotResiduals(growth.resids, form = factor(rep(1,nrow(growth))))) +
wrap_elements(~plotResiduals(growth.resids, quantreg = FALSE))+
wrap_elements(~testDispersion(growth.resids))

testUniformity(growth.resids)
plotResiduals(growth.resids, form = factor(rep(1,nrow(growth))))
plotResiduals(growth.resids, quantreg = FALSE)
testDispersion(growth.resids)
```

Summarise model
back transform from logit scale: by exponentiating on the odds scale. comparing to 1
```{r}
growth.brm3 |>
  summarise_draws(
    median,
    HDInterval::hdi,
    Pl = ~mean(.x < 0),
    Pg = ~mean(.x > 0),
    rhat,
    ess_bulk
    ) |>
  knitr::kable()
```

Probability scale
```{r}
growth.brm3 |>
  emmeans(~RegionCross) |>
  pairs()
```

Planned contrast:
CentralxCentral
Hybrid
NorthxNorth

```{r}
cmat=cbind(
  "Central vs North" = c(-1, 0, 1),
  "Central vs Hybrid" = c(-1, 1, 0),
  "North vs Hybrid" = c(0, 1, -1)
)
```


odds scale:
```{r}
growth.em<- growth.brm3 |>
  emmeans(~RegionCross) |>
  contrast(method=list(cmat)) |>
  gather_emmeans_draws() |>
#on odds ratio scale
  dplyr::summarise(median_hdci(.value),
            Pl = mean(.value < 0),
            Pg = mean(.value > 0)
            )
growth.em
```

#Plot
get variables
```{r}
growth.brm3 %>% get_variables()
```

```{r}
growth.plot<- growth.brm3 %>%
  emmeans(~RegionCross) %>%
  gather_emmeans_draws() %>%
    ggplot() +
  #geom_vline(xintercept = 0, linetype = "dashed") +
  stat_slab(aes(
    x = .value, y = RegionCross,
    fill = stat(ggdist::cut_cdf_qi(cdf,
      .width = c(0.5, 0.8, 0.95),
      labels = scales::percent_format()
    ))
  ), color = "black") +
  scale_fill_brewer("Interval", direction = -1, na.translate = FALSE) +
  theme_classic()+
  labs(x = "Growth (%)", y = "Population Cross")
growth.plot
```

```{r}
#saveRDS(growth.plot, file = "data/juv.growth.plot.region.rds")
growth.plot <- readRDS("data/juv.growth.plot.region.rds")
```
