---
title: "08_juvenile_colour_stats"
author: "Alex Macadam"
date: "2023-11-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##colour statistics
```{r}
library(tidyverse) #for data wrangling
library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(ggfortify) #for model diagnostics
library(sjPlot)    #for outputs
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(MuMIn)     #for AICc
library(brms)
library(broom.mixed)
library(tidybayes)
library(bayesplot)
#library(standist)   #for visualizing distributions
library(rstanarm)
library(ggeffects)
library(rstan)
library(DHARMa)
library(ggridges)
library(easystats)     #framework for stats, modelling and visualisation
library(patchwork)
source('/Users/alexmacadam/Documents/GitHub/AGF-Outplanting/helperFunctions.R')
```

```{r}
#load data
colour <- read.csv("/Users/alexmacadam/Documents/GitHub/AGF-Outplanting/data/juvenile_colour.csv", header=T, fileEncoding="UTF-8-BOM") |>
    mutate(Family = factor(Family),
         Cone = factor(Cone),
         Cross = factor(Cross),
         popcross = factor(popcross, levels = c('PIxPI', 'DAxDA', 'DAxAR', 'DAxMA', 'MAxDA', 'MAxMA', 'WDxMA')),
         RegionCross = factor(RegionCross),
         Breed = factor(Breed),
         Rack = factor(Rack),
         DeviceNo = factor(DeviceNo)
         ) |>
  filter(ColEndCor != "NA")
```

```{r}
colour<- colour |>
  mutate(ColEndCor2 = (100-(ColEndCor / 255)*100))
```

Check all entries are unique
```{r}
colour %>%
  dplyr::group_by(Cross, Rack, DevicePos, DeviceNo, IndID, popcross) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::filter(n > 1L)
```

```{r}
#colour<- colour %>% filter(Timepoint %in% c("T0","T1")) %>% dplyr::select(Family, Timepoint, popcross, Area, IndID, Cone, RegionCross, Rack, DeviceNo) %>%
#  pivot_wider(names_from = Timepoint, values_from = Area) %>%
#  tidyr::unnest() %>%
#  mutate(colour = ((`T1` - `T0`)/`T0`)*100) %>%
#  drop_na(colour)
```

# Fit the model
form
```{r}
colour.form <- bf(ColEndCor2 ~ popcross + (1|Rack),
                     family = gaussian()
                   )
```

```{r}
#colour |>
#group_by(popcross) |>
#summarise(
#mean(colour),
#median(colour),
#sd(colour),
#mad(colour)
#)
```

priors
```{r fitModel2h1, results='markdown', eval=TRUE, mhidden=TRUE, cache=TRUE}
priors<- prior("normal(55, 5)", class = "Intercept") +
    prior("normal(0, 5)", class = "b") +
    prior("normal(0, 5)", class = "sd", group = "Rack") +
   # prior("normal(0, 130)", class = "sd", group = "Timepoint") +
    prior(student_t(3, 0, 5), class = 'sigma') +
    prior(cauchy(0,2.5), class = 'sd')
```

run model
```{r}
colour.brm2 <- brm(colour.form, 
                  data = colour,
                  prior = priors,
                  sample_prior = 'yes',
                  iter = 5000,
                  warmup = 2500,
                  chains = 3, cores = 3,
                  thin = 5,
                  control = list(adapt_delta = 0.99, max_treedepth = 20),
                  refresh = 0,
                  backend = "rstan"
                  )
```

```{r}
#saveRDS(colour.brm2, file = "data/juv.colour.brm2.rds")
colour.brm2 <- readRDS("data/juv.colour.brm2.rds")
```

conditional effects
```{r}
colour.brm2 %>%
    conditional_effects("popcross") %>%
    plot(points = TRUE)
```

```{r}
colour.brm2 %>% SUYR_prior_and_posterior()
```

```{r}
colour.brm2$fit |> stan_trace()
```

```{r}
colour.brm2$fit |> stan_ac()
```

```{r}
colour.brm2$fit |> stan_rhat()
```

```{r}
colour.brm2$fit |> stan_ess()
```

```{r}
colour.brm2 |> pp_check(type = 'dens_overlay', nsamples = 250)
```

Dharma residuals
```{r}
colour.resids<- make_brms_dharma_res(colour.brm2, integerResponse = FALSE)
wrap_elements(~testUniformity(colour.resids)) +
wrap_elements(~plotResiduals(colour.resids, form = factor(rep(1,nrow(colour))))) +
wrap_elements(~plotResiduals(colour.resids, quantreg = FALSE))+
wrap_elements(~testDispersion(colour.resids))

testUniformity(colour.resids)
plotResiduals(colour.resids, form = factor(rep(1,nrow(colour))))
plotResiduals(colour.resids, quantreg = FALSE)
testDispersion(colour.resids)
```

back transform from logit scale: by exponentiating on the odds scale. comparing to 1
```{r}
colour.brm2 |>
  summarise_draws(
    median,
    HDInterval::hdi,
    Pl = ~mean(.x < 0),
    Pg = ~mean(.x > 0),
    rhat,
    ess_bulk
    ) |>
  knitr::kable()
```

Probability scale
```{r}
colour.brm2 |>
  emmeans(~popcross) |>
  pairs()
```

Planned contrast:
PIxPI
DAxDA 
DAxAR     
DAxMA              
MAxDA         
MAxMA              
WDxMA   

```{r}
cmat=cbind(
  "Pure vs Hybrid" = c(-1/3, -1/3, 1/4, 1/4, 1/4, -1/3, 1/4),
  "PureCentral vs Hybrid" = c(-1/2, -1/2, 1/3, 1/3, 1/3, 0, 0),
  "PureCentral vs Hybrid 2" = c(-1/2, -1/2, 1/4, 1/4, 1/4, 0, 1/4),
  "PureNorth vs Hybrid" = c(0, 0, 1/3, 1/3, 1/3, -1, 0),
  "PureNorth vs Hybrid2" = c(0, 0, 1/3, 1/3, 1/3, -1/2, -1/2),
  "Palms vs Davies" = c(-1, 1, 0, 0, 0, 0, 0),
  "MARmom vs MARdad" = c(0, 0, 0, 1, -1, 0, 0),
  "PureCentral vs PureNorth" = c(-1/2, -1/2, 0, 0, 0, 1/2, 1/2),
  "PureCentral vs PureNorth2" = c(-1/2, -1/2, 0, 0, 0, 1, 0)
)
```

odds scale:
```{r}
colour.brm2 |>
  emmeans(~popcross) |>
  contrast(method=list(cmat)) |>
  gather_emmeans_draws() |>
  summarise(median_hdci(.value),
            Pl = mean(.value < 0),
            Pg = mean(.value > 0)
            )
```
  "Pure vs Hybrid": y = -3.212. Hybrids have a 3.2 scale lower grey scale than pure juveniles (Pl = 1.000)
  
  "PureCentral vs Hybrid": y = -4.200. Hybrids have a 4.2 scale lower grey scale than pure central juveniles (Pl = 1.000)
  "PureCentral vs Hybrid 2": y = -3.594. Hybrids have a 3.6 scale lower grey scale than pure central juveniles (Pl = 1.000) 
  
  "PureNorth vs Hybrid": y = -3.093. Hybrids have a 3.1 scale lower grey scale than pure north juveniles (Pl = 0.995)
  "PureNorth vs Hybrid2": y = -2.718. Hybrids have a 2.7 scale lower grey scale than pure north juveniles (Pl = 0.987)
  
  "Palms vs Davies": y = 0.906. There is no evidence that Palms has different grey scale to davies (Pg = 0.761)
  
  "PureCentral vs PureNorth": y = -1.491. There is weak evidence that PureNorth have 1.5 scale lower grey scale than PureCentral juveniles (Pl = 0.934)
  "PureCentral vs PureNorth2": y = -1.085. There is no evidence that PureNorth have lower grey scale than PureCentral juveniles (Pg = 0.876)
  
  "MARmom vs MARdad": y = -4.918. Hybrids have a 4.9 scale lower grey scale than pure juveniles (Pl = 0.983)

#Plot
get variables
```{r}
colour.brm2 %>% get_variables()
```

```{r}
colour.plot<- colour.brm2 %>%
  emmeans(~popcross) %>%
  gather_emmeans_draws() %>%
    ggplot() +
  #geom_vline(xintercept = 0, linetype = "dashed") +
  stat_slab(aes(
    x = .value, y = popcross,
    fill = stat(ggdist::cut_cdf_qi(cdf,
      .width = c(0.5, 0.8, 0.95),
      labels = scales::percent_format()
    ))
  ), color = "black") +
  scale_fill_brewer("Interval", direction = -1, na.translate = FALSE) +
  theme_classic()+
  labs(x = "Colour (Grey Scale)", y = "Population Cross")
colour.plot
```

```{r}
#saveRDS(colour.plot, file = "data/juv.colour.plot.reef.rds")
colour.plot <- readRDS("data/juv.colour.plot.reef.rds")
```

#############
# Fit the model
## Random intercept
form
```{r}
colour.form <- bf(ColEndCor2 ~ RegionCross + (1|Rack),
                     family = gaussian()
                   )
```

```{r}
#colour |>
#group_by(popcross) |>
#summarise(
#mean(colour),
#median(colour),
#sd(colour),
#mad(colour)
#)
```

priors
```{r fitModel2h1, results='markdown', eval=TRUE, mhidden=TRUE, cache=TRUE}
priors<- prior("normal(55, 5)", class = "Intercept") +
    prior("normal(0, 5)", class = "b") +
    prior("normal(0, 5)", class = "sd", group = "Rack") +
   # prior("normal(0, 130)", class = "sd", group = "Timepoint") +
    prior(student_t(3, 0, 5), class = 'sigma') +
    prior(cauchy(0,2.5), class = 'sd')
```

run model
```{r}
colour.brm3 <- brm(colour.form, 
                  data = colour,
                  prior = priors,
                  sample_prior = 'yes',
                  iter = 5000,
                  warmup = 2500,
                  chains = 3, cores = 3,
                  thin = 5,
                  control = list(adapt_delta = 0.99, max_treedepth = 20),
                  refresh = 0,
                  backend = "rstan"
                  )
```

```{r}
#saveRDS(colour.brm3, file = "data/juv.colour.brm3.rds")
colour.brm3 <- readRDS("data/juv.colour.brm3.rds")
```

conditional effects
```{r}
colour.brm3 %>%
    conditional_effects("RegionCross") %>%
    plot(points = TRUE)
```

```{r}
colour.brm3 %>% SUYR_prior_and_posterior()
```

```{r}
colour.brm3$fit |> stan_trace()
```

```{r}
colour.brm3$fit |> stan_ac()
```

```{r}
colour.brm3$fit |> stan_rhat()
```

```{r}
colour.brm3$fit |> stan_ess()
```

```{r}
colour.brm3 |> pp_check(type = 'dens_overlay', nsamples = 250)
```

Dharma residuals
```{r}
colour.resids<- make_brms_dharma_res(colour.brm3, integerResponse = FALSE)
wrap_elements(~testUniformity(colour.resids)) +
wrap_elements(~plotResiduals(colour.resids, form = factor(rep(1,nrow(colour))))) +
wrap_elements(~plotResiduals(colour.resids, quantreg = FALSE))+
wrap_elements(~testDispersion(colour.resids))

testUniformity(colour.resids)
plotResiduals(colour.resids, form = factor(rep(1,nrow(colour))))
plotResiduals(colour.resids, quantreg = FALSE)
testDispersion(colour.resids)
```

back transform from logit scale: by exponentiating on the odds scale. comparing to 1
```{r}
colour.brm3 |>
  summarise_draws(
    median,
    HDInterval::hdi,
    Pl = ~mean(.x < 0),
    Pg = ~mean(.x > 0),
    rhat,
    ess_bulk
    ) |>
  knitr::kable()
```

Probability scale
```{r}
colour.brm3 |>
  emmeans(~RegionCross) |>
  pairs()
```

Planned contrast:
 CentralxCentral
 Hybrid
 NorthxNorth

```{r}
cmat=cbind(
  "Central vs North" = c(-1, 0, 1),
  "Central vs Hybrid" = c(-1, 1, 0),
  "North vs Hybrid" = c(0, 1, -1)
)
```

odds scale:
```{r}
colour.brm3 |>
  emmeans(~RegionCross) |>
  contrast(method=list(cmat)) |>
  gather_emmeans_draws() |>
  dplyr::summarise(median_hdci(.value),
            Pl = mean(.value < 0),
            Pg = mean(.value > 0)
            )
```

#Plot
get variables
```{r}
colour.brm3 %>% get_variables()
```

```{r}
colour.plot2<- colour.brm3 %>%
  emmeans(~RegionCross) %>%
  gather_emmeans_draws() %>%
    ggplot() +
  #geom_vline(xintercept = 0, linetype = "dashed") +
  stat_slab(aes(
    x = .value, y = RegionCross,
    fill = stat(ggdist::cut_cdf_qi(cdf,
      .width = c(0.5, 0.8, 0.95),
      labels = scales::percent_format()
    ))
  ), color = "black") +
  scale_fill_brewer("Interval", direction = -1, na.translate = FALSE) +
  theme_classic()+
  labs(x = "Colour (Grey Scale)", y = "Population Cross")
colour.plot2
```

```{r}
#saveRDS(colour.plot2, file = "data/juv.colour.plot.region.rds")
colour.plot2 <- readRDS("data/juv.colour.plot.region.rds")
```
