---
title: "11_adult_ITS2"
author: "Alex Macadam"
date: "2023-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)

if (!require("pacman")) install.packages("pacman")
pacman::p_load("dplyr", "edgeR", "ggplot2", "MCMC.OTU", "pairwiseAdonis", "rcartocolor",
               "RColorBrewer", "Redmonder", "reshape2", "vegan")

pacman::p_load_gh("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")


if (!require("edgeR")){BiocManager::install("edgeR", update = FALSE) 
  library(edgeR)}

options("scipen" = 10)
```

```{r}
its2Seq = read.delim("C:/Users/amacadam/Dropbox/PhD/Chapter 5- Outplanting/Chapter 5 Data/adult ITS2/439_20231130T121104_DBV_20231130T184910.seqs.absolute.abund_only2.txt", header = TRUE, check.names = FALSE)
head(its2Seq)

its2MetaData = read.delim("C:/Users/amacadam/Dropbox/PhD/Chapter 5- Outplanting/Chapter 5 Data/adult ITS2/439_20231130T121104_DBV_20231130T184910.seqs.absolute.meta_only.txt", header = TRUE, check.names = FALSE)
head(its2MetaData)

#its2Seq = cbind(its2Seq[1], its2MetaData[,2:3], its2Seq[,c(2:length(its2Seq))])
#colnames(its2Seq)[3] = "reef"
#head(its2Seq)

its2Seq$reef = factor(its2Seq$reef, levels = c("Fantome", "Davies", "Arlington", "Moore", "Martin", "Wood"))
levels(its2Seq$reef)

#its2Seq$Sample_site = factor(its2Seq$Sample_site, levels(its2Seq$Sample_site)[c(4, 2, 3, 1)])
#its2Seq = its2Seq[order(its2Seq$Sample_site, its2Seq$reef), ]
#levels(its2Seq$Sample_site)

head(its2Seq)
```

Count number of reads
```{r}
sum(its2MetaData$raw_contigs)
sum(its2MetaData$post_qc_absolute_seqs)
sum(its2MetaData$post_med_absolute)
```

First we purge sequences and transpose the data to work with edgeR
```{r}
goods = purgeOutliers(its2Seq, count.columns = 4:length(its2Seq), otu.cut = 0.0001, sampleZcut = -5)
```

```{r}
its2SeqTransposed = t(goods[, 4:length(goods[1, ])])
its2SeqList = DGEList(counts = its2SeqTransposed)
head(its2SeqList$samples)
```

Now we can use TMM normalization in edgeR
```{r}
its2SeqNorm =  calcNormFactors(its2SeqList, method = "TMM")
head(its2SeqNorm$samples)
its2TMM = t(cpm(its2SeqNorm, normalized.lib.sizes = TRUE))
its2SeqNorm = cbind(its2Seq[,c(1:3)], its2TMM)
colnames(its2SeqNorm)[4] = "Clade C"

head(its2SeqNorm)
```

Calculation of ITS2 sequence relative abundances
Now we can calculate the relative abundance of each ITS2 sequences per sample. This will allow us to view the assemblages in a faceted barplot.
```{r}
colOrder = order(colSums(its2SeqNorm[4:length(its2SeqNorm[1,])]), decreasing = FALSE) + 3

its2SeqPerc = cbind(its2SeqNorm[,c(1:3)], its2SeqNorm[,c(colOrder)])

its2SeqPerc$sum = apply(its2SeqPerc[, c(4:length(its2SeqPerc[1,]))], 1, function(x) {
  sum(x, na.rm = T)
})

its2SeqPerc = cbind(its2SeqPerc[, c(1:3)], (its2SeqPerc[, c(4:(ncol(its2SeqPerc)-1))] 
                                                / its2SeqPerc$sum))
```

Now a quick sanity check. If this worked the sum of each row should = 100% (i.e. “1”).
```{r}
apply(its2SeqPerc[, c(4:(ncol(its2SeqPerc)))], 1, function(x) {
  sum(x, na.rm = T)
})
```

Everything adds up to 1, this is good! The code works.

I added an additional column to sort better for the stacked barplot. This was just a work around to get the facet_grid() function to play nice with our data. I added a coulumn “barPlotOrder” and for each population I filled in a series 1:n for each sample at each Site:Depth combo, so now there’s no large blank expanses on the plot.
```{r}
sampleCounts = plyr::count(its2SeqPerc, c('reef'))
meltedList = melt(lapply(sampleCounts$freq,function(x){c(1:x)}))
its2SeqPerc$barPlotOrder = meltedList$value
its2SeqPerc=its2SeqPerc[c(1,ncol(its2SeqPerc),2:(ncol(its2SeqPerc)-1))]

head(its2SeqPerc)

gssSeq = otuStack(its2SeqPerc, count.columns = c(5:length(its2SeqPerc[1, ])),
                  condition.columns = c(1:4))[1:6674,] # remove summ rows
```

```{r}
gssSeq<- read.csv("C:/Users/amacadam/Dropbox/PhD/Chapter 5- Outplanting/Chapter 5 Data/adult ITS2/gssSeq.csv", header=T, fileEncoding="UTF-8-BOM")
```

```{r}
levels(gssSeq$otu)

#levels(gssSeq$reef) = c("Fantome", "Davies", "Arlington", "Moore", "Martin", "Wood")

gssSeq$reef = factor(gssSeq$reef, levels=c("Fantome", "Davies", "Arlington", "Moore", "Martin", "Wood"))
levels(gssSeq$reef)
```

Consruct ITS2 sequence barplot
```{r}
colorCount = length(c(5:length(its2SeqPerc[1,])))
getPalette = colorRampPalette(redmonder.pal(8, "qPBI"), bias = 1.7)
```

```{r}
counts<- gssSeq |>
  group_by(reef, otu) |>
  dplyr::summarise(count)

otus<- gssSeq |>
  group_by(reef) |>
  dplyr::summarise(otu)
```

```{r}
its2SeqPlotA = ggplot(gssSeq, aes(x = barPlotOrder2, y = count, fill = factor(otu))) +
  geom_bar(position = "stack", stat = "identity", color = "black",
           size = 0.25) + 
  ylab("Proportion") +
  scale_fill_manual(values=rev(getPalette(colorCount)))+ 
  
  labs(fill = expression(paste(italic("ITS2"), " sequence"))) +
  guides(fill = guide_legend(ncol = 9, reverse = TRUE)) +
  facet_grid( ~ reef, scales = "free_x") + #faceting plots by Depth and Site
  theme_bw()

its2SeqPlot = its2SeqPlotA +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.y = element_text(color = "black", size = 12),
        axis.text.y = element_text(color = "black", size = 12),
        legend.position = "bottom",
        legend.justification = "left",
        legend.direction = "horizontal",
        legend.title = element_text(color = "black", size = 12, hjust = 0.5, angle = 90),
        legend.text = element_text(color = "black", size = 10),
        legend.key = element_blank(),
        legend.key.size = unit(0.4,"line"),
        legend.background = element_blank(),
        panel.border = element_blank(),
        panel.background = element_rect(fill = "white"),
        plot.background = element_blank(),
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12),
        strip.background = element_rect(fill = "white", size = 0.9)
  )

its2SeqPlot
```

ITS2 type profiles
We can now look at the ITS2 type profiles predicted by SymPortal.

Prepare ITS2 type profile data
Similar to what we did with the sequence data
```{r}
its2Profs = read.delim("C:/Users/amacadam/Dropbox/PhD/Chapter 5- Outplanting/Chapter 5 Data/adult ITS2/439_20231130T121104_DBV_20231130T184910.profiles.absolute.abund_and_meta2.txt", header = TRUE, check.names = FALSE)
head(its2Profs)

its2Profs = cbind(its2Profs[1], its2MetaData[,2:3], its2Profs[,c(2:length(its2Profs))])
colnames(its2Profs)[3] = "reef"
head(its2Profs)

its2Profs$reef = factor(its2Profs$reef, levels = c("Fantome", "Davies", "Arlington", "Moore", "Martin", "Wood"))
levels(its2Profs$reef)

#its2Profs$Sample_site = factor(its2Profs$Sample_site, levels(its2Profs$Sample_site)[c(4, 2, 3, 1)])
#its2Profs = its2Profs[order(its2Profs$Sample_site, its2Profs$reef), ]
#head(its2Profs)

sampleCounts = plyr::count(its2Profs, c('reef'))
meltedList = reshape2::melt(lapply(sampleCounts$freq,function(x){c(1:x)}))
its2Profs$barPlotOrder = meltedList$value
its2Profs=its2Profs[c(1,ncol(its2Profs),2:(ncol(its2Profs)-1))]

head(its2Profs)
```

```{r}
its2Profs<- read.csv("C:/Users/amacadam/Dropbox/PhD/Chapter 5- Outplanting/Chapter 5 Data/adult ITS2/its2Profs.csv", header=T, fileEncoding="UTF-8-BOM")
```

Normalization of ITS2 type profile reads
Similar to what we did with the ITS2 sequences, but we won’t purge any low abundance reads, since we only have 13 total ITS2 type profiles.
```{r}
its2ProfsTransposed = t(its2Profs[, 5:length(its2Profs[1, ])])
its2ProfsList = DGEList(counts = its2ProfsTransposed)
head(its2ProfsList$samples)

its2ProfsNorm =  calcNormFactors(its2ProfsList, method = "TMM")
head(its2ProfsNorm$samples)
its2TMM = t(cpm(its2ProfsNorm, normalized.lib.sizes = TRUE))
its2ProfsNorm = cbind(its2Profs[,c(1:4)], its2TMM)

head(its2ProfsNorm)
```

Preparing ITS2 type profiles for plotting
```{r}
colOrder2 = order(colSums(its2ProfsNorm[5:length(its2ProfsNorm[1,])]), decreasing = TRUE) + 4

its2ProfsPerc = cbind(its2ProfsNorm[,c(1:4)],its2ProfsNorm[,c(colOrder2)])
its2ProfsPerc$sum = apply(its2ProfsPerc[, c(5:length(its2ProfsPerc[1,]))], 1, function(x) {
  sum(x, na.rm = T)
})

its2ProfsPerc = cbind(its2ProfsPerc[, c(1:4)], (its2ProfsPerc[, c(5:(ncol(its2ProfsPerc)-1))] 
                                              / its2ProfsPerc$sum))
head(its2ProfsPerc)
```

```{r}
apply(its2ProfsPerc[, c(5:(ncol(its2ProfsPerc)))], 1, function(x) {
  sum(x, na.rm = T)
})
```

Again, everything looks good!
```{r}
gssProf = otuStack(its2ProfsPerc, count.columns = c(5:length(its2ProfsPerc[1, ])),
               condition.columns = c(1:4))[1:470,] # remove summ rows

levels(gssProf$otu)

levels(gssProf$reef)
levels(gssProf$reef) = c("Fantome", "Davies", "Arlington", "Moore", "Martin", "Wood")
levels(gssProf$reef)
```

Consruct ITS2 type profile barplot
```{r}
colorCount2 = length(c(4:length(its2ProfsPerc[1,]))) +1
getPalette2 = colorRampPalette(redmonder.pal(8, "qPBI"))
```

```{r}
its2ProfsPlotA = ggplot(gssProf, aes(x = barPlotOrder, y = count, fill = factor(otu))) +
  geom_bar(position = "stack", stat = "identity", color = "black", size = 0.25) + 
  ylab("Proportion") +
  scale_fill_manual(values = c(getPalette2(colorCount2)[2:12], 
                               carto_pal(n = 7, "Tropic")[c(6,7)]))+
  labs(fill = expression(paste(italic("ITS2"), " type profile"))) +
  guides(fill = guide_legend(ncol = 2, reverse = FALSE)) +
  facet_grid(~ reef, scales = "free_x") + #faceting plots by Depth and Site
  theme_bw()

its2ProfsPlot = its2ProfsPlotA +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.y = element_text(color = "black", size = 12),
        axis.text.y = element_text(color = "black", size = 12),
        legend.position = "bottom",
        legend.title = element_text(color = "black", size = 12, hjust = 0.5, angle = 90),
        legend.text = element_text(color = "black", size = 10),
        legend.key = element_blank(),
        legend.key.size = unit(0.75,"line"),
        legend.background = element_blank(),
        panel.border = element_blank(),
        panel.background = element_rect(fill = "white"),
        plot.background = element_blank(),
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12),
        strip.background = element_rect(fill = "white", size = 0.9)
  )

its2ProfsPlot
```

Statistical analyses on ITS2 type profiles
Cheking dispersion with PERMDISP
Using betadisper() in vegan to look at multivariate homogeneity of dispersion (PERMDISP) between sites. This is using Bray-Curtis dissimilarity.
```{r}
set.seed(694) #setting seed allows repetition of randomized processes

its2dispS = betadisper(vegdist(its2ProfsNorm[, c(5:ncol(its2ProfsNorm))]), its2ProfsNorm$reef)

anova(its2dispS)
```

significant effect of reef on betadiversity.

Permutation test for pairwise comparisons
Follow up with a permutation test to see where differences occur.
```{r}
set.seed(694)

its2PermTest = permutest(its2dispS, permutations = 9999, pairwise = T, model = "full", )
its2PermTest
```

```{r}
its2PermTest$statistic
```

Running PERMANOVA in R
Now let’s see how different communities are from each other with PERMANOVA. We will utilize the adonis() function in vegan. We will use Bray-Curtis similarity for our distance matrix and run a total 0f 9,999 permutations, and test the effects of reef. Dispersion is heteroschedastic, but PERMANOVA is robust to deviations in homgeneity of variance ( Anderson and Walsh, 2013)
```{r}
set.seed(694)
its2Adonis = adonis(its2ProfsNorm[, c(5:ncol(its2ProfsNorm))] ~ reef, 
data = its2ProfsNorm, permutations = 9999, method = "bray")

its2Adonis
```
PERMANOVA reveals that reef has a signifcant effect on Symbiodiniaceae associations in our A. hyacinthus samples.

Pairwise PERMANOVA for multiple comparisons
Since we found that Depth was a significant factor in our PERMANOVA we can now use pairwise PERMANOVA to reveal where differences occur across depth. This utilizes the package pairwiseAdonis, where we will again use Bray-Curtis similarity and 9,999 permutations. We also have added false discovery rate (FDR) corrections since we are perfoming multiple comparisons.
```{r}
set.seed(694)
its2PWAdonis = pairwise.adonis(its2ProfsNorm[, c(5:ncol(its2ProfsNorm))],
                               factors = its2ProfsNorm$reef,
                               sim.method = "bray", p.adjust.m = "BH", perm = 9999)
                
its2PWAdonis

#write.csv(its2PWAdonis, "C:/Users/amacadam/Dropbox/PhD/Chapter 5- Outplanting/Chapter 5 Data/adult ITS2/its2PWAdonis.csv")
```
We see that again see differences between samples.

SIMPER test between north and south populations
Similarity percentage test (SIMPER) will show us the ITS2 type profiles that contribute the most to the dissimilarity between depth zones
```{r}
depths = its2ProfsNorm$Depth_zone
levels(depths) = c("Shallow","Shallow","Deep","Deep")
depths
its2ProfsNorm$depth = depths
its2ProfsNorm = cbind(its2ProfsNorm[c(1:4)],its2ProfsNorm[ncol(its2ProfsNorm)],its2ProfsNorm[c(5:(ncol(its2ProfsNorm)-1))])
head(its2ProfsNorm)
its2SimperD = simper(sqrt(its2ProfsNorm[, c(6:ncol(its2ProfsNorm))]), its2ProfsNorm$depth)
summary(its2SimperD)
```
NMDS

ITS2 NMDS
```{r}
#make community matrix - extract columns with abundance information
com = its2SeqNorm[,4:ncol(its2SeqNorm)]
```

```{r}
#turn abundance data frame into a matrix
m_com = as.matrix(com)
```

```{r}
set.seed(123)
nmds = metaMDS(m_com, distance = "bray")
nmds
```

```{r}
plot(nmds)
```

```{r}
#extract NMDS scores (x and y coordinates) for sites from newer versions of vegan package
data.scores = as.data.frame(scores(nmds)$sites)
```

```{r}
#add columns to data frame 
data.scores$name = its2SeqNorm$name
data.scores$reef = its2SeqNorm$reef
#data.scores$Type = its2SeqNorm$Type
 
head(data.scores)
```

```{r}
library(ggplot2)

xx = ggplot(data.scores, aes(x = NMDS1, y = NMDS2)) + 
    geom_point(size = 4, aes(colour = reef))+ 
    theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), 
    axis.text.x = element_text(colour = "black", face = "bold", size = 12), 
    legend.text = element_text(size = 12, face ="bold", colour ="black"), 
    legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), 
    axis.title.x = element_text(face = "bold", size = 14, colour = "black"), 
    legend.title = element_text(size = 14, colour = "black", face = "bold"), 
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2),
    legend.key=element_blank()) + 
    labs(x = "NMDS1", colour = "Reef", y = "NMDS2", shape = "Type")  + 
    scale_colour_manual(values = c("#264653","#287271", "#2A9D8F", "#8AB17D", "#E9C46A", "#F4A261", "#E76F51")) 
 
xx
```

```{r}
library(patchwork)

ITS2_plot<- its2SeqPlot / its2ProfsPlot
```

```{r}
ggsave("C:/Users/amacadam/Dropbox/PhD/Chapter 5- Outplanting/Chapter 5 Figures/Mock plots/Figure_4_ITS2.pdf", ITS2_plot, device="pdf", width=8, height=8, units = "in")
```

```{r}
ggsave("C:/Users/amacadam/Dropbox/PhD/Chapter 5- Outplanting/Chapter 5 Figures/Mock plots/Figure_4_NMDS.pdf", xx, device="pdf", width=6, height=4, units = "in")
```